Concurrent Programming with CXL

### 2020-06-05 book/paper.tex
Working on trying to identify blocking vs busy waiting
Working on reader/writer lock

CXL is heavily influenced by TLA+, TLC, and PlusCal~\cite{}.
CXL is designed to have a lower learning curve than those
systems, but is not as powerful.  When you finish this book
and want to learn more, we strongly encourage checking
those out.

@@

\chapter{Reader/Writer Locks}

@@

While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}:
processes spin in a loop until some desirable condition is met.
The astute reader might wonder if obtaining an ordinary lock itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the lock is available
(see Figure~\ref{fig:locks}).

In most operating systems and programming language runtimes, however,
when a process acquires a lock, the process is placed on a scheduling
queue and stops using CPU cycles until the lock becomes available.
Moreover, if there are multiple processes waiting for a lock, there
is a scheduler that decides which of the processes gets the lock.
Busy waiting disables all this.

@@

@@

So it is considered ok to have processes using spinlock a implement a
lock, but not to busy-wait for application-specific conditions.

@@

>>>>

CXL is heavily influenced by Leslie Lamport's work on
TLA+, TLC, and PlusCal~\cite{}.
CXL is designed to have a lower learning curve than those
systems, but is not as powerful.  When you finish this book
and want to learn more, we strongly encourage checking
those out.
Another excellent resource is Fred Schneider's book ``On
Concurrent Programming.''

@@

\chapter{Reader/Writer Locks using Busy Waiting}

@@

While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}:
processes spin in a loop until some desirable application-level condition is met.

The astute reader might wonder if obtaining an ordinary lock itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the lock is available
(see Figure~\ref{fig:locks}).
In most operating systems and programming language runtimes, however,
when a process acquires a lock, the process is placed on a scheduling
queue and stops using CPU cycles until the lock becomes available.
Moreover, if there are multiple processes waiting for a lock, there
is a scheduler that decides which of the processes gets the lock.
Busy waiting disables all this.

@@

We call a process \emph{blocked} in a state if steps by that process
do not change the shared memory in that state.  A process trying to
acquire a test-and-set spinlock held by another process is a good example
of a process being blocked: the process only reads the shared memory
but does not change it.

However, a process trying to obtain a read or write lock in
Figure~ref{fig:rwbusy} has to actually obtain a spinlock just to check
if the read or write lock is available.  So even when the read or write
lock is not available, the process ends up updating the spinlock and
therefore the shared memory.

Thus, it is considered ok to have processes be blocked while waiting
for application-specific conditions, but not for them to
busy-wait for application-specific conditions.

[TODO.  The following must be updated.]
CXL does not complain about Figure~\ref{fig:rwbusy}, but using
the right test, CXL can be used to check for busy-waiting.
One approach is as follows.  First, obtain a lock in the initialization
code by adding the following line to the end of the code:
\begin{code}
\begin{verbatim}
call acquire_rlock();
\end{verbatim}
\end{code}

Second, run CXL with the \texttt{-b} flag.  This flag tells CXL to check
that all states are non-terminating and have all processes blocked.

\chapter{Reader/Writer Lock, Part 2}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWmulti.cxl}
\end{code}
\caption{Checking that multiple readers can acquire the read lock.}
\label{fig:rwmulti}
\end{figure}

@@

While CXL does not detect any issues, we must remember what it is checking
for: it checks to make sure that there are never a reader and a writer
in the critical section, and there are never two readers in the critical
section.  Moreover, it checks progress: all executions eventually terminate.
What it does \emph{not} check is if the code allows more than one reader
in the critical section.  Indeed, we could have implemented the reader/writer
lock as in Figure~\ref{fig:rwmulti} using an ordinary lock, never allowing
more than one process in the critical section.

Figure~\ref{fig:rwmulti} contains a new test that ensures that indeed more
than one reader can enter the critical section at a time.  It does so by
having processes sometimes wait in the critical section until all other readers
are there.  If other readers cannot enter, then that process enters in an
infinite loop that CXL will readily detect.  Try it out!

An important lesson here is that one should not celebrate too early if CXL
does not find any issues.  While a test may be successful, a test may not
explore all desirable executions.  While CXL can help explore all cornercases
in a test, it cannot find problems that the test is not looking for.
